Is the direction of Granger causal influence the same as the direction of information flow?

### 2015-07-10 paper.tex
Edit equation to fit within margins.
Make extreme edits to fit equations in margins.

Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$. Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,weissman}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.

@@

We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison. However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).

@@

Through this scheme, the estimate $\widehat{\Theta}_{i}$ is seen to converge to $\Theta$ in mean-square sense as $i\rightarrow\infty$:
\begin{align*}
	\mathbb{E}[\widehat{\Theta}_{i}] &= \mathbb{E}\bigg[\Theta+\frac{1}{i}\sum_{j=1}^{i}N_{j}\bigg] = \mathbb{E}[\Theta]+0 \\
	\mathbb{E}[(\widehat{\Theta}_{i}-\Theta)^{2}] &= \mathbb{E}\bigg[\bigg(\frac{1}{i}\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] \\
												  &= \frac{1}{i^{2}}\mathbb{E}\bigg[\bigg(\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] = \frac{i}{i^{2}}\mathbb{E}[N_{1}^{2}] = \frac{\sigma_{N}^{2}}{i}\overset{i\to\infty}\longrightarrow 0.
\end{align*}

@@

In the reverse direction,
\begin{align*}
	I(0*\widehat\Theta^{n-1} & \rightarrow X^n) = \sum_{i=0}^{n-1}{I(X_{i+1}; \widehat\Theta^i | X^i)} \\
							 &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis \label{eq:dir-info-rev-noisy} \\
							 &= \frac{1}{2} \log \bigg( 2 \pi e \frac{\sigma_N^2 + \sigma_R^2}{\sigma_R^2} \bigg) \\
							 & \; \; + \sum_{i=2}^{n-1} \bigg( \frac{1}{2} \log \bigg( 2 \pi e \text{Var} \bigg[ R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg] \bigg) \\
							 & \qquad \qquad - \frac{1}{2} \log( 2 \pi e \sigma_R^2 ) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:dir-info-rev-noisy}.

@@

For a given time instant $t$, the $i$\textsuperscript{th} trial is modeled as
\begin{equation*}
	X_t^{(i)} = \sum_{j=1}^p \alpha_j(t) X_{t-j}^{(i)} + \epsilon_t^{(i)}
\end{equation*}
Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
\begin{align*}
	\left[ \begin{array}{c}
			X_t^{(1)} \\
			X_t^{(2)} \\
			\vdots \\
			X_t^{(N)}
	\end{array} \right]
	&=
	\left[ \begin{array}{ccc}
			X_{t-1}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			X_{t-1}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			\vdots & \ddots & \vdots \\
			X_{t-1}^{(N)} & \cdots & X_{t-p}^{(N)} \\
			%X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			%X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			%\vdots &  & \ddots & \vdots \\
			%X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
	\end{array} \right]
	\left[ \begin{array}{c}
			\alpha_1(t) \\
			\alpha_2(t) \\
			\vdots \\
			\alpha_p(t)
	\end{array} \right] +
	%&\qquad \qquad \qquad +
	\left[ \begin{array}{c}
			\epsilon_t^{(1)} \\
			\epsilon_t^{(2)} \\
			\vdots \\
			\epsilon_t^{(N)}
	\end{array} \right]
\end{align*}

@@

						& \qquad             \bigg( \frac{1}{q-1} \sum_{k=1}^{q-2} R_k - \frac{1}{q-1} \sum_{k=1}^{q-1} N_k - R_{q-1} \bigg) \bigg] \\
					    &= \frac{\min\{p-2, q-2\}}{(p-1)(q-1)} \sigma_R^2 + \frac{\min\{p-1, q-1\}}{(p-1)(q-1)} \sigma_N^2 \\

@@

	\mathbb{E}[X_1 X_j] &= \mathbb{E} \bigg[ \Theta \bigg( \frac{1}{j-1} \sum_{k=1}^{j-2} R_k - \frac{1}{j-1} \sum_{k=1}^{j-1} N_k - R_{j-1} \bigg) \bigg] = 0 \\

>>>>

Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$. Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,jiao2013universal}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.

@@

We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison. However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,hesse2003use,jiao2013universal}).

@@

Through this scheme, the estimate $\widehat{\Theta}_{i}$ is seen to converge to $\Theta$ in mean-square sense as $i\rightarrow\infty$:
\begin{align*}
	\mathbb{E}[\widehat{\Theta}_{i}] &= \mathbb{E}\bigg[\Theta+\frac{1}{i}\sum_{j=1}^{i}N_{j}\bigg] = \mathbb{E}[\Theta]+0 \\
	\mathbb{E}[(\widehat{\Theta}_{i}-\Theta)^{2}] &= \mathbb{E}\bigg[\bigg(\frac{1}{i}\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] \\
												  &= \frac{1}{i^{2}}\mathbb{E}\bigg[\bigg(\sum_{j=1}^{i}N_{j}\bigg)^{2}\bigg] \\
												  &= \frac{i}{i^{2}}\mathbb{E}[N_{1}^{2}] = \frac{\sigma_{N}^{2}}{i}\overset{i\to\infty}\longrightarrow 0.
\end{align*}

@@

In the reverse direction,
\begin{align*}
	I(0*\widehat\Theta^{n-1} & \rightarrow X^n) = \sum_{i=0}^{n-1}{I(X_{i+1}; \widehat\Theta^i | X^i)} \\
							 &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis \label{eq:dir-info-rev-noisy} \\
							 &= \frac{1}{2} \log \bigg( 2 \pi e \frac{\sigma_N^2 + \sigma_R^2}{\sigma_R^2} \bigg) \\
							 & + \sum_{i=2}^{n-1} \bigg( \frac{1}{2} \log \bigg( 2 \pi e \text{Var} \bigg[ R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg] \bigg) \\
							 & \qquad \qquad - \frac{1}{2} \log( 2 \pi e \sigma_R^2 ) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:dir-info-rev-noisy}.

@@

For a given time instant $t$, the $i$\textsuperscript{th} trial is modeled as
\begin{equation*}
	X_t^{(i)} = \sum_{j=1}^p \alpha_j(t) X_{t-j}^{(i)} + \epsilon_t^{(i)}
\end{equation*}
Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
\begin{align*}
	\left[ \!\! \begin{array}{c}
			X_t^{(1)} \\
			X_t^{(2)} \\
			\vdots \\
			X_t^{(N)}
	\end{array} \!\! \right]
	&=
	\left[ \! \begin{array}{ccc}
			X_{t-1}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			X_{t-1}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			\vdots & \ddots & \vdots \\
			X_{t-1}^{(N)} & \cdots & X_{t-p}^{(N)} \\
			%X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			%X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			%\vdots &  & \ddots & \vdots \\
			%X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
	\end{array} \! \right]
	\left[ \!\! \begin{array}{c}
			\alpha_1(t) \\
			\alpha_2(t) \\
			\vdots \\
			\alpha_p(t)
	\end{array} \!\! \right] +
	%&\qquad \qquad \qquad +
	\left[ \!\! \begin{array}{c}
			\epsilon_t^{(1)} \\
			\epsilon_t^{(2)} \\
			\vdots \\
			\epsilon_t^{(N)}
	\end{array} \!\! \right]
\end{align*}

@@

						& \qquad             \bigg( \frac{1}{q-1} \sum_{k=1}^{q-2} R_k - \frac{1}{q-1} \sum_{k=1}^{q-1} N_k - R_{q-1} \bigg) \bigg]
\vspace{-12pt}
					    &= \frac{\min\{p-2, q-2\}}{(p-1)(q-1)} \sigma_R^2 \\
						& \qquad + \frac{\min\{p-1, q-1\}}{(p-1)(q-1)} \sigma_N^2 \\

@@

	\mathbb{E}[X_1 X_j] &= \mathbb{E} \bigg[ \Theta \bigg( \frac{1}{j-1} \sum_{k=1}^{j-2} R_k \\
						& \qquad \qquad \qquad - \frac{1}{j-1} \sum_{k=1}^{j-1} N_k - R_{j-1} \bigg) \bigg] = 0 \\
